---
title: "Three-Pointers in the NBA"
author: "Gop Arop"
format: revealjs
---

## Data
 
* Question of interest investigated: How prominent has the 3 point shot been in the last 24 years?

* Data sources: HoopR & BasketballAnalyzeR packages, BasketballReference.com

* Packages and scraping contains thousands of observations on team/indivudal statistics. 

* Stats can come in the form of per game, totals, or even in-game for a specific game.

## Visualization 1

```{r, warning=FALSE,echo=FALSE}
library(tidyverse)
library(hoopR)
library(BasketballAnalyzeR)
shot_data <-  tibble(nba_shotchartdetail(league_id = '00', player_id = '2544', season = "2023-24"))

data <- shot_data[[1]][[1]] |> 
  mutate(LOC_X = as.numeric(LOC_X), LOC_Y = as.numeric(LOC_Y), xx = LOC_X/10, yy = LOC_Y/10 - 41.75) |> 
  rename(shot_distance = SHOT_DISTANCE, original_x = LOC_X, original_y = LOC_Y, converted_y = yy, converted_x = xx)
```

```{r}
chart <- ggplot(data = data, aes(x = converted_x,y = converted_y)) + geom_point() + coord_fixed() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_rect(fill = "white"), axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank())

drawNBAcourt(chart)
```


## Visualization 2

When you click the **Render** button a presentation will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```
